Epistemic Semantics in Guarded String Models\thanks{Thanks guys.}

### 2021-01-15 tex/pdraft.tex
rewrite introduction

\begin{document}
\maketitle
\begin{abstract}
  Constructive and computable multi-agent epistemic possible worlds
  models are defined, where possible worlds models are guarded string
  models in an epistemic extension of Kleene Algebra with Tests.  The
  account is framed as a formal language Epik (Epistemic \textsc{Kat}) for
  defining such models.  The language is implemented by translation
  into the finite state calculus, and alternatively by modeling
  propositions as lazy lists in Haskell. The syntax-semantics
  interface for a fragment of English is defined by a categorial
  grammar.
\end{abstract}

@@

\section{Introduction and Related Work}
Linguistic semantics in the Montague tradition proceeds by assigning
propositional {\em semantic values} to disambiguated sentences of a
natural language.  A proposition is a set or class of {\em possible
  worlds}.  These are often assumed to be things with the same nature
and complexity as the world we occupy \citep{lewis1986plurality}.  But
alternatively, one can work with small idealized models, in order to
illustrate and test ideas.  The point of this paper is to scale up toy
or idealized models to countable sets of worlds, and to constructive
and computable modeling of epistemic alternatives for agents.  We
describe a systematic way of defining such models, and
illustrate how to apply them in natural language semantics.  The focus
is on epistemic semantics and clausal embedding. The fundamental move
is to identify possible worlds with strings of primitive events, so
that propositions are sets of strings. An advantage in this is that it
allows for a mathematical description of an algebra of propositions,
coupled with a computational representation using either lazy lists of
strings, or finite state machines that describe sets of strings.

The approach taken here synthesizes five antecedents.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}.  {\em Action models} in dynamic
epistemic semantics introduce the technique of constructing epistemic
models from primitive alternative relations on events, in order to
capture the epistemic consequences of perceptual and communicative
events \citep{baltag1999logic}.  Literature on {\em finite state
  methods in linguistic semantics} has used event strings and sets of
event strings to theorize about tense and aspect in natural language
semantics
\citep{fernando2004finite,fernando2007observing,carlson2009tense} and
to express intensions \cite{fernando2017intensions}.  Work on {\em
  finite state intensional semantics} has investigated how to do the
semantics of intensional complementation in a setting where
compositional semantics is expressed in a finite state calculus
\citep{rooth2017finite,collard2018finite}.

A running example of an event-sequence model is {\em The Concealed Coin}. Amy
and Bob are seated at a table. There is a coin on the table under a cup, heads
up. The coin could be heads-up $H$ or tails-up $T$, and neither agent knows
which it is. This initial situation is possible world $w_1$. We can construct
additional worlds by sequencing events like those shown in \ex{coin1}.

@@

The worlds $w_1,\ldots,w_4$ in \ex{coin2} are examples constructed from the
events in \ex{coin1}. Juxtaposition indicates the order and occurence of events,
for example $w_3$ indicates that after the events of $w_2$, event $b_1$ occurs,
fully, this indicates beginning in world $w_1$, Amy surreptitiously peeking at
a heads-up coin, and then Bob doing the same.

@@

The truth values for English sentences are observed at the end of the event
sequences specified by $w_1,\ldots, w_4$, where $0$ stands for falsity and $1$
for truth.

@@

The events come with pre-conditions and post-conditions.  Amy can turn
the coin from heads to tails only if the coin is heads-up, so $\atHT$
has the pre-condition of the coin being heads up. Once she turns the
coin over, tails must be face-up, so $a_{10}$ has the post-condition
of the coin being tails-up. Let $h$ be the Boolean proposition that
the coin is heads up and $t$ be the Boolean proposition that the coin
is tails-up.  Then pre- and post-conditions can be described by
Boolean formulas, with $h$ being the pre-condition of $\atHT$ and
$t$ being the post-condition. This is expressed using an operator
``$:$'' (read ``and next'') that pairs Boolean formulas. The formula
$\ev h t$ describes $\atHT$ (Amy turning the coin from heads
to tails) as happening only in an $h$ state, and concluding in a
$t$ state.  Events don't have to change state: the event $\apT$
(Amy peeking at tails) can happen only in a $t$ state, and does
not change the state ($\ev{t}{t}$).

@@

However a coin cannot be showing both heads and tails! Currently the
precondition $h$ of $\atHT$ only says that heads must be showing, and
says nothing about the fact that tails must be face-down, indicated by
the formula $\bar t$. We will further restrict the feasible conditions
for our actions by restricting the space of valuations for our formulas.

A sequence such as $\bar{h}{t}$ can be viewed both as a formula and as
a valuation of primitive propositions, which we use to describe world
state. The primitives are listed in fixed order, and left unmarked
(indicating true) or marked with the overbar (indicating false).
Since a coin is heads or tails but not both, we want to allow the
valuations $h\bar{t}$ and $\bar{h}t$, and disallow $ht$ and
$\bar{h}\bar{t}$.\footnote{Two Booleans constrained by a state formula, rather than one Boolean, are stipulated here to
 illustrate state formulas.
}  This is enforced by a {\em state formula}, which is
a Boolean formula, in this case the one given on the second line of
\ex{effect}.  Where {\sf{B}} is a set of state primitives and $\varphi$ is a
 constraint over {\sf{B}},
${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\varphi}$ is the set of valuations of
{\sf{B}} that make formula $\varphi$ true. Valuations are called atoms,
because they correspond to the atoms of a Boolean algebra of tests
\citep{kozen2001automata}.

Formulas that describe pre- and
post-conditions are {\em effect formulas}. They are interpreted as
defining relations between atoms, as defined in Figure
\ref{effectfigure}.  The atoms they relate are constrained by the
state formula as well.  For the heads-tails example, let the state
formula and the effect formula for $a_1$ (Amy peeking at heads) be as
specified in \ex{effect}. Then ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\varphi}$
and the relation on atoms for the event $a_1$ are as given at
the bottom in \ex{effect}.

\begin{example}   \label{effect}
\(\begin{array}[t]{ll}
{\sf{B}} & \{h,t\}\\
\mbox{state formula $\varphi$} & h\bar{t} + \bar{h}t \\
\mbox{effect formula $\zeta$ for $a_1$} &  h \mathbin{:} h \\
 \mathcal{A}^\varphi_{\mathsf B} & \{h\bar{t},\bar{h}t\}   \\
\sem{\zeta}^\varphi & \{\pair{h\bar{t}}{h\bar{t}}\}   \\
\end{array}\)
\end{example}

@@

\section{Epistemic guarded string models}

@@

\section{The logical language of Epistemic \textsc{Kat}}
The standard language for Kleene algebra with tests has the signature
$\tuple{K,+,\cdot,*,\bar{},0,1}$ \citep{kozen2001automata}.  In a guarded string
model for \textsc{Kat}, $K$ is a set of sets of guarded strings, $+$ is set union, the
operation $\cdot$ is fusion product raised to sets, $*$ is Kleene star, the operation
$\;\bar{}\;$ is complement for tests, $0$ is the empty set, and $1$ is the set of
atoms.\footnote{$0$ has the dual role the identity for $+$ (union), and as False for
  operations on tests.  $1$ has the dual role of the identity for product (fusion
  product raised to sets), and True for tests.  } To this we add a unary
modal operation $\Diamond_a$ for each agent, and a unary complement operation $\neg$
on elements of $K$.  Intuitively, $\Diamond_a p$ is the set of
worlds where proposition $p$ is epistemically possible for agent $a$.  Propositional
complement is included because natural languages have sentence negation.  In
addition, universal box modalities are defined as duals of existential diamond
modalities.

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and
regular relations between strings that was designed for use in
computationonal phonology and morphographemics
\citep{kaplan1994regular,beesley2003finite}.  Current implementations
allow for the definition of functions on regular sets and relations
\citep{hulden2009foma,linden2009hfst,karttunen2010update}.  Such definitions are used
here to construct of a model for epistemic \textsc{Kat} inside the finite state
calculus (Figures~\ref{fstfigure1}\&\ref{relclosure}).

@@

\section{Bounded Lazy Interpretation}

@@

Epik is closely related to Kleene Algebra with Hypotheses
\textsc{Ka+H}~\cite{cohen1994hypotheses}, which permits user-specification of
futher equations beyond the axioms of \textsc{Ka}. In practice, these hypotheses
are used to express domain-specific properties of the events, such as the
commutativity. We can model both the state formulae and the event formulae as
\textsc{KA+H} hypotheses. Specifically, the state formula $\varphi$ can be
thought of as a hypothesis $\varphi \equiv 1$, and an event formula $\zeta_a$
for event $a$, is simply the hypothesis $a \equiv \sum_{\varphi,\varphi' \in
  \llbracket \zeta_a \rrbracket} \varphi;a;\varphi'$. In general, the Horn
theory of \textsc{KA+H} is undecidable, however many special cases
are~\cite{doumane2019kleene, hardin2002elimination,
  kozen1997kleenecompleteness}.

The key facet that differentiates Epik from \textsc{Ka+H} is the modality. Each
alternative relation represents a set of uni-directional rewrites that must be
applied for that agent, rather than the universal, optional rewrites represented
by the \textsc{Ka+H} hypotheses. It may be possible to encode Epik's alternative
relations into \textsc{Ka+H} hypotheses, but such an encoding is future work.

>>>>

\begin{document}
\maketitle
\begin{abstract}
  Constructive and computable multi-agent epistemic possible worlds models are
  as sets of guarded string models in an epistemic extension of Kleene Algebra
  with Tests (\textsc{Kat}). The account is framed as a formal language Epik
  (Epistemic \textsc{Kat}) for defining such models. The language is implemented
  by translation into the finite state calculus, and alternatively by modeling
  propositions as lazy lists in Haskell. The syntax-semantics interface for a
  fragment of English is defined by a categorial grammar.
\end{abstract}

@@

\section{Introduction and Related Work}
Linguistic semantics in the Montague tradition proceeds by assigning
propositional {\em semantic values} to disambiguated sentences of a natural
language. A proposition is a set or class of {\em possible worlds}, which are
often assumed to have the same nature and complexity as the world we occupy
\citep{lewis1986plurality}. But alternatively, one can work with small idealized
models, to illustrate and test ideas. The point of this paper is to extend
standard idealized models to countable sets of worlds, and to constructively and
computably model alternatives for epistemic agents.

Enter Epik, which is a systematic way of defining such models, and can be
applied to natural language semantics, specifically, epistemic semantics and
clausal embedding. The fundamental insight is to identify possible worlds with
strings of primitive events, so that propositions are sets of (guarded) strings,
whose regular sets have a rich algebraic characterization~\cite{kozen1997kleene}
and computational model~\cite{kozen1997kleenecompleteness,kozen2001automata}.
Epik leverages these algebrae to define a mathematical model
(Sections~\ref{sec:model} \&~\ref{sec:logic}) and computationally interpret
using finite state machines (Section~\ref{sec:fst}) and lazy lists of strings
(Section~\ref{sec:lazy})).

{\em Related Work.} Epik synthesizes five antecedents.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}.  {\em Action models} in dynamic
epistemic semantics introduce the technique of constructing epistemic
models from primitive alternative relations on events, in order to
capture the epistemic consequences of perceptual and communicative
events \citep{baltag1999logic}.  Literature on {\em finite state
  methods in linguistic semantics} has used event strings and sets of
event strings to theorize about tense and aspect in natural language
semantics
\citep{fernando2004finite,fernando2007observing,carlson2009tense} and
to express intensions \cite{fernando2017intensions}.  Work on {\em
  finite state intensional semantics} has investigated how to do the
semantics of intensional complementation in a setting where
compositional semantics is expressed in a finite state calculus
\citep{rooth2017finite,collard2018finite}.

{\em Example.} Consider an example event-sequence model called {\em The
  Concealed Coin}. Amy and Bob are seated at a table. There is a coin on the
table under a cup, heads up. The coin could be heads-up $H$ or tails-up $T$, and
neither agent knows which it is. Call this initial possible world $w_1$.
Additional worlds can be constructed via the events in \ex{coin1}.

@@

The worlds in \ex{coin2} are examples constructed from the events in \ex{coin1}.
Juxtaposition indicates the order of events; for example, $w_3$ indicates that
after the events of $w_2$, event $b_1$ occurs, which corresponds to, in the
initial world, Amy surreptitiously peeking at a heads-up coin, and then Bob
doing the same.

@@

Now, the truth ($1$) or falsity ($0$) of English sentences can be evaluated in
each world $w_i$, i.e. \emph{after} the events have happened. See \ex{coin3} for
examples, such as in $w_4$, where Bob knows it's heads (because he peeked after
turning the coin), but Amy does not (because Bob secretly turned the coin over).

@@

To create realistic models, contradictory event sequences must be prohibited.
For example, $\apH\atTH$ would require the coin to simultaneously heads-up and
tails-up. To mechanize this, events in Epik come with Boolean pre-conditions and
post-conditions. In the running example, the Boolean variables $h$ (the coin is
heads-up) and $t$ (the coin is tails-up) represent the coin's state. Then, since
Amy can only observe heads when the coin is heads-up, event $\apH$ has the
invariant condition $h$. Similarly, Amy can only turn the coin from tails to
heads if it shows tails, so $\atTH$ has the tails-up precondition $t$ and the
heads-up postcondition $h$.

Pre- and post- conditions are expressed using an operator ``$:$'' (read ``and
next'') that pairs Boolean formulas into an \emph{effect formula}. Specifically,
the effects for $\apH$ are written $\ev h h$, and for $\atTH$ are written $\ev t
h$. Effect formulae are interpreted relations on boolean valuations, as defined
in Figure~\ref{effectfigure}.

However, a coin cannot simultanously show both heads and tails! Currently the
precondition $h$ of $\atHT$ only says that heads must be showing, and says
nothing about the fact that tails must be face-down, indicated by the formula
$\bar t$. The effect valuations are restricted to only the feasible ones via a
\emph{state formula}, shown in \ref{effect}, and demonstrated for $\apH$ in
\ref{effect}. The state formula, also shown in \ref{effect}, says that the coin
is either heads-up or tails-up, but not both\footnote{Two Booleans constrained
  by a state formula, is a more illustrative example than using a single boolean
  and modelling tails as $\bar h$.}. Here juxtaposition represents conjuction,
$+$ is disjunction, and overbar is negation.

With the state formula in hand, $\apH\atTH$ is evidently contradictory, the
post-condition $h$ of $\apH$ is incompatible with the pre-condition $t$ of
$\atTH$. See Figure~\ref{effectfigure} for more details.

@@

Given a set {\sf{B}} of state primitives and $\varphi$ a state formula over
{\sf{B}}, define ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\varphi}$ to be the set of
valuations of {\sf{B}} that make formula $\varphi$ true. The valuations for the
coin example are shown in \ref{effect}, using the sequence notation for
valuations, e.g. $\bar{h}{t}$, wherein every state primitive is listed in fixed
order, and left unmarked (indicating true) or marked with the overbar
(indicating false). Valuations are called atoms, because they correspond to the
atoms of a Boolean algebra of tests \citep{kozen2001automata}.

\begin{example}   \label{effect}
\(\begin{array}[t]{ll}
{\sf{B}} & \{h,t\}\\
\mbox{state formula $\varphi$} & h\bar{t} + \bar{h}t \\
\mbox{effect formula $\zeta_{a_1}$ for $a_1$} &  h \mathbin{:} h \\
 \mathcal{A}^\varphi_{\mathsf B} & \{h\bar{t},\bar{h}t\}   \\
\sem{\zeta_{a_1}}^\varphi & \{\pair{h\bar{t}}{h\bar{t}}\}   \\
\end{array}\)
\end{example}

Thus far, the presentation is closely related to Kleene Algebra with Hypotheses
(\textsc{Ka+H})~\cite{cohen1994hypotheses}, which permits user-specification of
futher equations beyond the axioms of \textsc{Ka}. Specifically, the state
formula $\varphi$ can be thought of as a hypothesis $\varphi \equiv 1$, and an
event formula $\zeta_a$ for event $a$, is simply the hypothesis $a \equiv
\sum_{\varphi,\varphi' \in \llbracket \zeta_a \rrbracket} \varphi;a;\varphi'$.
Rather than rely on general techniques ~\cite{doumane2019kleene,
  hardin2002elimination, kozen1997kleenecompleteness}, we relalize our specific
hypotheses directly.

The key facet that differentiates Epik from \textsc{Ka+H} is the epistemic
modality, which cannot be naively represented using universal
hypotheses\footnote{The existence of this encoding is an interesting future
  topic}. Sections~\ref{sec:model} and~\ref{sec:logic} discuss the
model-theoretic and logical development of the epistemic operators.

@@

\section{Epistemic guarded string models}
\label{sec:model}

@@

\section{The logical language of Epistemic \textsc{Kat}}
\label{sec:logic}

The standard language for Kleene algebra with tests has the signature
$\tuple{K,+,\cdot,*,\bar{},0,1}$ \citep{kozen2001automata}.  In a guarded string
model for \textsc{Kat}, $K$ is a set of sets of guarded strings, $+$ is set union, the
operation $\cdot$ is fusion product raised to sets, $*$ is Kleene star, the operation
$\;\bar{}\;$ is complement for tests, $0$ is the empty set, and $1$ is the set of
atoms.\footnote{$0$ has the dual role the identity for $+$ (union), and as False for
  operations on tests.  $1$ has the dual role of the identity for product (fusion
  product raised to sets), and True for tests.  } To this we add a unary
modal operation $\Diamond_a$ for each agent, and a unary complement operation $\neg$
on elements of $K$.  Intuitively, $\Diamond_a p$ is the set of
worlds where proposition $p$ is epistemically possible for agent $a$.  Propositional
complement is included because natural languages have sentence negation.  In
addition, universal box modalities are defined as duals of existential diamond
modalities.

@@

\section{Translation into the finite state calculus}
\label{sec:fst}

The finite state calculus is an algebra of regular sets of strings and
regular relations between strings that was designed for use in
computationonal phonology and morphographemics
\citep{kaplan1994regular,beesley2003finite}.  Current implementations
allow for the definition of functions on regular sets and relations
\citep{hulden2009foma,linden2009hfst,karttunen2010update}.  Such definitions are used
here to construct of a model for epistemic \textsc{Kat} inside the finite state
calculus (Figures~\ref{fstfigure1}\&\ref{relclosure}).

@@

\section{Bounded Lazy Interpretation}
\label{sec:lazy}

@@
